---
title: "Combining taxonomies"
output: html_notebook
---

If there is one thing we believe in, are the classifications curated by Ana and Belén from Porifera, Chordata and Echinodermata.



```{r}
ASV.last <- read_csv(here("data", "new_ASV_after_swarm.csv"))

Hash.clas.last <- read_csv(here("data", "Hash_ids_NorthAtlantic.csv"))

taxonomy.cols <- colnames(Hash.clas.last[-1])

Hash.clas.last |> 
  mutate(precision = case_when(species == "NA" & genus == "NA" & family !="NA" ~ "family",
                                species == "NA" & genus != "NA" & family !="NA" ~ "genus",
                                species != "NA" & genus != "NA" & family !="NA" ~ "species")) |> 
  semi_join(ASV.last) |>
  select(Hash, precision) -> precision
 Hash.clas.last |> 
  unite (-Hash, col = "taxa", sep = "%") -> Hash.clas.last
 
 Hashes.Seqs <- read_csv(here("data/hash_key_220.csv"))
```


```{r}
Hash.clas.last |> 
   ungroup() |> 
   filter (str_detect(taxa, "Chordata")) |> 
   distinct(taxa) |> arrange(taxa) |> separate(taxa, into = taxonomy.cols, sep = "%", remove = F)

# Callionymiformes vs Syngnathiformes

Hash.clas.last |> 
  mutate(taxa = str_replace(taxa, "Syngnathiformes", "Callionymiformes")) -> Hash.clas.last
# Helicolenus appears duplicated
 Hash.clas.last |> 
   filter (str_detect(taxa, "Helicolenus")) |> slice(1) |> pull(taxa) -> Helicolenus

  Hash.clas.last |> 
    filter (str_detect(taxa, "Helicolenus")) |> 
    mutate(taxa = Helicolenus) -> Helicolenus
# Lycodes has conflicting taxonomy  
   Hash.clas.last |> 
    filter (str_detect(taxa, "Lycodes")) |> 
     separate(taxa, into = taxonomy.cols, sep = "%") |> 
     mutate(order = "Perciformes") |> 
     unite(-Hash, col = "taxa", sep = "%")-> Lycodes
# Scomber appears duplicated
  Hash.clas.last |> 
    filter (str_detect(taxa, "Scomber")) |> 
    mutate(taxa = "Metazoa%Chordata%Actinopteri%Scombriformes%Scombridae%Scomber%Scomber scombrus") -> Scomber

#   
  
## Revisit NAs from Echinodermata
  
  Hash.clas.last |> 
    filter (str_detect(taxa, "Echinodermata")) |> 
    filter (str_detect(taxa, "NA")) |> 
    left_join(Hashes.Seqs) |> 
    eDNAfuns::fasta.writer(sequence = Sequence, header = Hash, file.out = here("data", "Echinodermata_na.fasta"))

  ## Revisit NAs from Porifera
  
  Hash.clas.last |> 
    filter (str_detect(taxa, "Porifera")) |> 
     filter (str_detect(taxa, "NA")) |> 
    left_join(Hashes.Seqs) |> 
    eDNAfuns::fasta.writer(sequence = Sequence, header = Hash, file.out = here("data", "Porifera_na.fasta"))
  
 Hash.clas.last |> 
   inner_join(ASV.last) |> 
   
   
    filter (str_detect(taxa, "Porifera")) |> 
     filter (str_detect(taxa, "Geodia")) |> 
   filter (!str_detect(taxa, "barretti|hentscheli|parva")) |> 
   group_by(Hash, taxa ) |> 
   summarise(nr = sum(nReads)) |> 
   left_join(Hashes.Seqs) |> 
   mutate(len = str_length(Sequence)) |> 
   arrange(taxa) |> 
   separate(taxa, into = taxonomy.cols, sep = "%") |> 
   mutate(species = str_replace_all(species, " ", "_"), 
            header = paste0(str_sub(Hash, 1, 10),";organism=",species)) |> 
    eDNAfuns::fasta.writer(sequence = Sequence, header = header, file.out = here("data", "Porifera_geodia.fasta")) 
  
  ## Run them in bold and compare results
  
    bolds <- read_csv2(here("data/BOLD/parsed_echino_porifera.csv"))
    
   Hash.clas.last |> 
    filter (str_detect(taxa, "Echinodermata|Porifera")) |> 
    filter (str_detect(taxa, "NA")) |> 
     left_join(bolds) |> 
     mutate (precision.bold = case_when(is.na(species) & is.na(genus) & !is.na(family) ~ "family",
                                        is.na(species) & !is.na(genus) & !is.na(family) ~ "genus",
                                        !is.na(species) & !is.na(genus) & !is.na(family) ~ "species")) |> 
     unite(kingdom, phylum, class, order, family, genus, species,col = "taxa.bold", sep = "%") |> 
     separate(taxa, into = taxonomy.cols, sep = "%", remove = F) |> 
     mutate (precision.blast = case_when(species == "NA" & genus == "NA" & family !="NA" ~ "family",
                                species == "NA" & genus != "NA" & family !="NA" ~ "genus",
                                species != "NA" & genus != "NA" & family !="NA" ~ "species")) |> 
     mutate(final.id = case_when(precision.bold == "species" ~ taxa.bold,
                                 precision.bold == "genus" ~ taxa.bold,
                                 precision.blast== "family"~ taxa,
                                 precision.bold == "family"~ taxa.bold,
                                 TRUE                      ~ taxa)) |> 
     select(Hash, taxa = final.id) -> Echino_and_porifera

```


```{r}

  Helicolenus |> 
  bind_rows(Lycodes) |> 
  bind_rows(Scomber) |> 
  bind_rows(Echino_and_porifera) -> improved
 # nothing shared between efforts



Hash.clas.last |> 
  anti_join(improved, by = "Hash") |> 
  bind_rows(improved) -> Hash.clas.last

## What THings have we blasted

last.blast <- read_csv(here("pipeline_output","hash_classified_2023_97_100.csv"))

### Keep the improved

miscellaneous.bold <- read_csv(here("data", "BOLD", "miscellaneous.bold.csv"))

miscellaneous.bold |> 
  semi_join(improved, by = "Hash")

## HAsh clas, keep only CHordata, Ehcino, porifera

Hash.clas.last |> 
  ungroup() |> 
   filter (str_detect(taxa, "Chordata|Echinodermata|Porifera")) -> goods

last.blast |> 
  anti_join(goods, by = "Hash") |> 
  unite(-Hash, col = "taxa", sep = "%") |> 
  bind_rows(goods) -> almost_there 
  
# What to do with the bolds

almost_there |> 
anti_join(miscellaneous.bold, by = "Hash") |> 
  bind_rows(miscellaneous.bold) -> done
  
  
  
ASV.last |> group_by(Hash) |> summarise ( nr = sum(nReads) ) |> inner_join(Hashes.Seqs) |> mutate(len = str_length(Sequence)) |> inner_join(done)  |>
  group_by(str_detect(taxa, "NA|NA|NA|NA|NA|NA") ) |> summarise(n(), sum(nr))

ASV.last |> group_by(Hash) |> summarise ( nr = sum(nReads) ) |> inner_join(Hashes.Seqs) |> mutate(len = str_length(Sequence)) |> anti_join(done)  |> summarise(n(), sum(nr))


```
Geodia sp. is actually two OTUs: most of them are G. barretti, there is a different one with 304 bp

```{r}
done |> 
  filter (str_detect(taxa, "Geodia sp.$")) |> 
  separate(taxa, into = taxonomy.cols, sep = "%") |> 
  mutate(species = case_when(Hash == "ca1f1d837d711a86099cf35047ef511cc30d64b4"~ "Geodia sp. 2 PC-2011",
                             TRUE                                              ~ "Geodia barretti") ) |> 
  select(Hash, taxonomy.cols) |> 
  unite(-Hash, col = "taxa", sep = "%") -> gsp
 
```

Geodia sp. 2 are actually a whole bunch of things. Two of them are only one bp apart, and we can consider them the same thing. 
BLAST puts them in between phelagrei, vaubani y parva. BOLD has them to phelagrei.     

```{r}
tricky.geodias <- read_csv2(here("data/BOLD/parsed_porifera.csv"))

tricky.geodias |> 
  replace_na(list (species= "sp.")) |> 
  unite(genus, species, col = "species", sep = " ", remove = F) |> 
  select(Hash, taxonomy.cols) |> 
  unite(-Hash, col = "taxa", sep = "%") |> 
  slice (1:14) -> gsp2

gsp |> 
  bind_rows(gsp2) %>%
  anti_join(done,., by= "Hash") |> 
  bind_rows(gsp) |> 
  bind_rows(gsp2) -> done_and_dusted

## We forgot to rejoin with the Hash.clas.last

Hash.clas.last |> 
  unite(taxonomy.cols, col = "taxa", sep = "%") |> 
  anti_join(done_and_dusted, by = "Hash") |> 
  bind_rows(done_and_dusted) |> 
  write_csv(here("data/final_id_2023_on_Friday.csv"))
```

# Update on aug 2023

For Chordata, Cnidaria, Porifera and Echinodermata, if we have a genus but reached no conclussion on species, the species field would be genus sp.

```{r}
taxonomies <- read_csv(here("data/final_id_2023_on_Friday.csv"))

taxonomies |> 
  separate(taxa, into = taxonomy.cols, sep = "%") |> 
  filter (phylum %in% c("Chordata","Echinodermata", "Porifera", "Cnidaria")) |> 
  mutate (species = case_when(species != "NA" ~ species,
                              genus != "NA"   ~ paste0(genus, " sp."),
                              genus == "NA"   ~ species)) |> 
  unite (taxonomy.cols, col = "taxa", sep = "%")-> updated.tax

taxonomies |> 
  anti_join(updated.tax, by = "Hash") |> 
  bind_rows(updated.tax) |> write_csv(here("data/final_id_Aug_2023.csv"))

```

## Appartently we are not done yet

There are some instances in which we have done order sp. or family sp.

```{r}
taxonomies <- read_csv(here("data/final_id_Aug_2023.csv"))

taxonomies |> 
  separate(taxa, into = taxonomy.cols, sep = "%") |>  
  filter (genus == "NA" & str_detect(species, "sp."))

taxonomies |> 
  mutate (taxa = case_when(str_detect(taxa, "Poraniidae%NA")    ~ "Metazoa%Echinodermata%Asteroidea%Valvatida%Poraniidae%Porania%Porania sp.",
                           str_detect(Hash, "10cba565eed241d964cdcef6ae7ec2334433ed4e") ~ "Metazoa%Porifera%Demospongiae%Tetractinellida%Geodiidae%Geodia%Geodia sp. 1",
                            str_detect(Hash, "ca1f1d837d711a86099cf35047ef511cc30d64b4") ~ "Metazoa%Porifera%Demospongiae%Tetractinellida%Geodiidae%Geodia%Geodia sp. 2",
                           str_detect(taxa, "Pennatulidae%NA%NA")~ "Metazoa%Cnidaria%Anthozoa%Pennatulacea%Pennatulidae%Ptilella%Ptilella grayi",
                           str_detect(taxa, "Pennatulacea%NA%NA%NA")~"Metazoa%Cnidaria%Anthozoa%Scleralcyonacea%Funiculinidae%Funiculina%Funiculina quadrangularis",
                            str_detect(taxa, "Funiculina")      ~"Metazoa%Cnidaria%Anthozoa%Scleralcyonacea%Funiculinidae%Funiculina%Funiculina quadrangularis",
                           str_detect(taxa, "Hygrosoma")        ~ "Metazoa%Echinodermata%Echinoidea%Echinothurioida%Echinothuriidae%Hygrosoma%Hygrosoma sp.",
                           str_detect(taxa, "Phyrella")         ~ "Metazoa%Echinodermata%Holothuroidea%Dendrochirotida%Phyllophoridae%Molpadia%Molpadia borealis",
                           str_detect(taxa, "Astropecten")      ~ "Metazoa%Echinodermata%Asteroidea%Paxillosida%Astropectinidae%Astropecten%Astropecten irregularis",
                           str_detect(taxa, "Phyllophoridae%NA")~ "Metazoa%Echinodermata%Holothuroidea%Dendrochirotida%Phyllophoridae%Pentamera%Pentamera calcigera",
                           str_detect(taxa, "Rossellidae%NA")   ~ "Metazoa%Porifera%Hexactinellida%Lyssacinosida%Rossellidae%Caulophacus%Caulophacus arcticus",
                           str_detect(Hash, "bb0da6fd980e9935e8ec148f209bfaaa0d62fac1")   ~ "Metazoa%Porifera%Demospongiae%Poecilosclerida%Hymedesmiidae%Hymedesmia%Hymedesmia primitiva",
                           str_detect(taxa, "Niphatidae")       ~ "Metazoa%Porifera%Demospongiae%Haplosclerida%Niphatidae%NA%Niphatidae sp.",
                           str_detect(taxa, "Clytia sp.")       ~ "Metazoa%Cnidaria%Hydrozoa%Leptothecata%Campanulariidae%Clytia%Clytia sp.",
                           str_detect(Hash, "baeef57023d47979237aa78cbd47e50b468abb70") ~ "Metazoa%Cnidaria%Anthozoa%Alcyonacea%Isididae%Acanella%Acanella arbuscula",
                           str_detect(Hash, "77bfcabb7ed93ce22017b19f41c8b7692a0a5852") ~ "Metazoa%Cnidaria%Anthozoa%Alcyonacea%Alcyoniidae%Anthomastus%Anthomastus grandiflorus",
                           str_detect(Hash, "d3d38edc5742e249f9fe56a5a427e5f2fd02a830") ~ "Metazoa%Cnidaria%Anthozoa%Actiniaria%Actinostolidae%Maractis%Maractis sp.",
                           str_detect(Hash, "bd45a6fefa56c886f2a7a45d03c1a270bdb40caa") ~ "Metazoa%Cnidaria%Anthozoa%Actiniaria%Actinostolidae%Actinostola%Actinostola sp.",
      TRUE                                 ~ taxa
                           )) -> taxonomies
  # filter (str_detect(taxa, "Poraniidae%NA")) |> 
  # mutate (taxa = "Metazoa%Echinodermata%Asteroidea%Valvatida%Poraniidae%Porania%Porania sp.") -> Porania

# taxonomies |> 
#   filter (str_detect(taxa, "Hygrosoma")) |> 
#   mutate (taxa = "Metazoa%Echinodermata%Echinoidea%Echinothurioida%Echinothuriidae%Hygrosoma%Hygrosoma sp.")-> Hygrosoma

# taxonomies |> 
#   filter (str_detect(taxa, "Phyrella"))|> 
#   mutate (taxa = "Metazoa%Echinodermata%Holothuroidea%Dendrochirotida%Phyllophoridae%Molpadia%Molpadia borealis")-> Molpadia

# taxonomies |> 
#   filter (str_detect(taxa, "Astropecten")) |> 
#   mutate (taxa = "Metazoa%Echinodermata%Asteroidea%Paxillosida%Astropectinidae%Astropecten%Astropecten irregularis") -> Astropecten

# taxonomies |> 
#   filter (str_detect(taxa, "Phyllophoridae%NA")) |> 
#   mutate (taxa = "Metazoa%Echinodermata%Holothuroidea%Dendrochirotida%Phyllophoridae%Pentamera%Pentamera calcigera")-> Pentamera

# Porifera
# 
# taxonomies |> 
#   filter (str_detect(taxa, "Rossellidae%NA")) |> 
#    mutate (taxa = "Metazoa%Porifera%Hexactinellida%Lyssacinosida%Rossellidae%Caulophacus%Caulophacus arcticus")-> Pentamera

# taxonomies |> 
#   filter (str_detect(Hash, "bb0da6fd980e9935e8ec148f209bfaaa0d62fac1")) |> 
#   mutate (taxa = "Metazoa%Porifera%Demospongiae%Poecilosclerida%Hymedesmiidae%Hymedesmia%Hymedesmia primitiva")-> Pentamera

# taxonomies |> 
#   separate(taxa, into = taxonomy.cols, sep = "%") |> 
#   filter(family != "NA" & genus == "NA" & species == "NA") |> 
#   filter (kingdom == "Metazoa") |> View()
 
  


taxonomies |> 
  filter (str_detect(taxa, "Cnidaria")) |> 
  separate(taxa, into = taxonomy.cols, sep = "%") |> 
  filter(genus == "NA" & species == "NA" & family != "NA") |> 
  filter (family %in% c("Sphaerocorynidae","Gorgoniidae", "Oceaniidae", "Actiniidae", "Pandeidae", "Gonactiniidae" )) |> 
  mutate (species = paste0(family , " sp.")) |> 
  unite (taxonomy.cols, col = "taxa", sep = "%") -> news

taxonomies |> 
  anti_join(news, by = "Hash") |> 
  bind_rows(news) -> taxonomies
  
taxonomies |> 
  filter (str_detect(taxa, "Cnidaria")) |> 
  separate(taxa, into = taxonomy.cols, sep = "%") |> 
  filter(genus == "NA" & species == "NA" & order != "NA" ) |> 
  semi_join(ASV.last) |> 
  mutate (species = paste0(order , " sp.")) |> 
  unite (taxonomy.cols, col = "taxa", sep = "%") -> news

# news |> 
# inner_join(Hashes.Seqs) |> 
#   fasta.writer(sequence = Sequence, header = Hash, file.out = here("nofam.fa"))
# 

taxonomies |> 
  anti_join(news, by = "Hash") |> 
  bind_rows(news) -> taxonomies

# taxonomies |> 
#   filter (str_detect(taxa, "Pennat")) |> 
#   filter (str_detect(taxa, "NA")) |> 
#   inner_join(Hashes.Seqs) |> 
#   fasta.writer(sequence = Sequence, header = Hash, file.out = here("Pennatulaceas.fa"))
# 
# taxonomies |> 
#   filter (str_detect(taxa, "Penna")) |> 
#   inner_join(ASV.last)
# 
# 
taxonomies |>
  filter (str_detect(Hash, "bd45a6fefa56c886f2a7a45d03c1a270bdb40caa")) |>
  inner_join(ASV.last)

taxonomies |>
  filter (str_detect(taxa, "Clytia sp.")) |>
  inner_join(ASV.last)
```
```{r}
taxonomies |> 
  write_csv(here("data/taxonomies_2023_08_18.csv"))
```

