---
title: "Refining taxonomical assignments"
output: html_notebook
editor_options: 
  chunk_output_type: console
---
Some problems with the taxonomical IDs

Take the last iteration of the files.

Add the info we have gotten from BOLD

For those with low quality matches, translate the sequences and make a blast of those translations - this will give us an accurate if not very precise taxonomical information

```{r}
library(here)
library(tidyverse)
library(taxonomizr)
library(insect)

ASV.last <- read_csv(here("data", "new_ASV_after_swarm.csv"))
Hash.clas.last <- read_csv(here("data", "Hash_ids_NorthAtlantic.csv"))
taxonomy.cols <- colnames(Hash.clas.last[-1])
Hash.clas.last |> 
  mutate(precision = case_when(species == "NA" & genus == "NA" & family !="NA" ~ "family",
                                species == "NA" & genus != "NA" & family !="NA" ~ "genus",
                                species != "NA" & genus != "NA" & family !="NA" ~ "species")) |> 
  semi_join(ASV.last) |> select(Hash, precision) -> precision
 Hash.clas.last |> 
  unite (-Hash, col = "taxa", sep = "%") -> Hash.clas.last 

Hash.clas.last |> 
   semi_join(ASV.last) |> 
 left_join(ids_thresholds[[2]] %>% 
  dplyr::rename(Hash =1)) |> 
  # filter (str_detect(taxa, "Chordata")) |> 
  distinct(taxa, consensus) |> View()
# 





Metadata.last  <- read_csv(here("data/metadata_north_atlanctic.csv")) |> 
  mutate (Host_name = case_when (Host == "Gb" ~"Geodia barretti",
                                 Host == "Gp" ~ "Geodia parva",
                                 Host == "Gh" ~"Geodia hentscheli",
                                 Host == "Pv" ~"Phakellia ventilabrum"))

Hashes.Seqs <- read_csv(here("data/hash_key_220.csv"))

taxonomy <- read_rds(here("data", "taxonomy.rds"))

lineages <- read_rds(here("data", "lineages_2023_07_03.rds"))

```

Find all the data from BOLD

```{r}
bolds <- list.files(path = here("data/BOLD"), full.names = T) |> map(~read_delim(.x, delim = "\t", col_types = c("cccii"))) |> bind_rows() |> separate (`Search DB`, into = c("SearchDB", "bestID", "worstID"), sep = "\t")

bolds |> 
  # filter (`Best ID` != "No match") |> 
  select (1:5) |> 
  select (Hash = 1, name = 2 ) |> 
  separate(Hash, into = "Hash", sep = ";") -> Bold_ids

Bold_ids |> 
  left_join(taxonomy) |> 
  distinct(taxID) |> 
  filter (!is.na(taxID)) |> 
  pull()  %>% 
  set_names(., nm =.) |> 
  map(~get_lineage(.x, db = taxonomy)) |> 
  map_dfr(~ bind_cols(.x) |>  mutate (rank = names(.x)) ,.id = "taxID") -> lineages.bold


lineages.bold |> 
     filter (rank %in% c( "kingdom", "phylum", "class", "order", "family","genus" , "species")) |> 
  pivot_wider(names_from = rank, values_from = 2) |> 
  mutate(taxID = as.numeric(taxID)) |> 
  right_join(Bold_ids |> 
              left_join(taxonomy) |> 
              select(Hash,taxID )) |> 
  select(-taxID) |> 
  unite(-Hash, col = "taxa", sep = "|") |> 
  select(Hash, taxa) -> newIDs.BOLD

newIDs.BOLD |> write_csv(here("data", "BOLD", "miscellaneous.bold.csv"))
  
```
BLASTx searches

```{r}
output <- read_delim(here("data/blastx/output.txt"), 
     delim = "\t", escape_double = FALSE, 
     col_names = c("qseqid", "sseqid",  "pident", "length" ,"mismatch", "gapopen" ,"qstart","qend", "sstart", "send" ,"evalue" ,"bitscore", "staxid", "qlen"), trim_ws = TRUE)

# Bring the previous lineages we used in this project
# 
# lineages1<- read_rds(here("data/lineages_2023-04-21.rds"))
# lineages2<- read_rds(here("data/lineages_2023-04-19.rds"))
# lineages3 <- read_csv(here("data/taxonomies_matches_today.csv"))
# 
# 
# # lineages 3 has almost all the info
# 
# output |> 
#   filter (pident > 90) |> 
#   separate_rows(staxid, sep = ";", convert = T) |> 
#   distinct(staxid) |> 
#   dplyr::rename(taxID = 1) |> 
#   anti_join(lineages) |> 
#   pull() |> 
#   map(insect::get_lineage, taxonomy) -> new.lineages
# 
# new.lineages <- set_names(x = new.lineages, nm = output |> 
#   filter (pident > 90) |> 
#   separate_rows(staxid, sep = ";", convert = T) |> 
#   distinct(staxid) |> 
#   dplyr::rename(taxID = 1) |> 
#   anti_join(lineages3) |> 
#   pull())
# 
# new.lineages %>% discard(function(x) is.na(x[[1]]) ) ->  new.lineages.good
# 
# new.lineages.good|> 
#   map_dfr(~ bind_cols(.x) |>  mutate (rank = names(.x)) ,.id = "taxID") -> new.lineages.good
# 
# new.lineages.good |> 
#   dplyr::rename(name = 2) |> 
#   filter (rank %in%  c( "kingdom", "phylum", "class", "order", "family","genus" , "species")) |> 
#   mutate(taxID = as.numeric(taxID)) -> new.lineages.good
# 
# new.lineages.good |> 
#   bind_rows(lineages3)|> 
#   filter (rank %in%  c( "kingdom", "phylum", "class", "order", "family","genus" , "species")) |> 
#   distinct() -> all.lineages


```


Join Output and lineages

```{r}
custom.lca <- function (df, cutoff = 90, keep.unresolved = F) {  # this function allows to change cutoff parameters for a specified dataframe (df)
    # ADDED on APR 2023
    # IF there are matches with the structure FAMILY sp. RAndmom
    # KEep them only if there aren't equally good matches with full details
  if (!isTRUE(keep.unresolved)){
    df |> 
      filter (!str_detect(species, " sp.")) -> df
  }  
  df %>%  
  group_by(qseqid) %>%
  select( pident, kingdom, phylum, class, order, family, genus, species) %>%
  nest() %>% # for each query, calculate the agreed taxonomy
  # ungroup %>% slice (1:10) %>%
  mutate(consensus = purrr::map(data,  function(.x) {
    # If there are 100% matches - keep those and calculate the LCA
   
    if (max(.x$pident == 100 )){
       .x %>%
        filter(pident == 100) %>%
        select(-pident) %>%
        condenseTaxa() %>% # agreement in Phylogeny
      paste(., collapse = "%")
      
    }else{
       # If there are no 100% matches, then keep things better than our cutoff
    if(max(.x$pident > cutoff )){

      .x %>%
        filter(pident > cutoff) %>%
        select(-pident) %>%
        condenseTaxa() %>% # agreement in Phylogeny
      paste(., collapse = "%")

       

    }else{
      # If there are no matches, better than the cutoff, then keep everything above 90
      
    .x %>%
         filter(pident > 90) %>%
        select(-pident) %>%
    condenseTaxa() %>%
       paste(., collapse = "%")
      }
  }
  }
  
  # Collapse all the taxa data separatated by %, como queda feo para leer en excel lo modificamos con # PERO es un lio dejarlo sin el % porq deja la table separada sin heads asi que mejor dejarlo como esta y luego en R separar las columnas por % y asignarles nombres
  
  )) %>%
  select(qseqid, consensus) %>%
  unnest(consensus)}
```

```{r}
lineages |> 
  group_by(taxID, rank) |> 
  tally() |> 
  filter (n >1) |> 
  distinct(taxID) %>% 
  anti_join(lineages,.) |> 
  pivot_wider(names_from = "rank", values_from = "name") %>%
  left_join(output |> 
              filter (pident > 90) |>
              separate_rows(staxid, sep = ";", convert = T) |> 
              dplyr::rename(taxID = staxid),.) -> ready


```

Do it

```{r}
custom.lca(df = ready,cutoff = 97,keep.unresolved = F) -> finalid_blastx

finalid_blastx |> 
  separate(qseqid, into = "qseqid", sep = ";size=") -> finalid_blastx
```

## BLASTP with the best translation

```{r}
output_blastp <- read_delim(here("data/blastp/output_2023_06_30.txt"), 
     delim = "\t", escape_double = FALSE, 
     col_names = c("qseqid", "sseqid",  "pident", "length" ,"mismatch", "gapopen" ,"qstart","qend", "sstart", "send" ,"evalue" ,"bitscore", "staxid", "qlen"), trim_ws = TRUE)

# output_blastp |>  filter (pident > 90) |> 
#   separate_rows(staxid, sep = ";", convert = T) |> 
#   distinct(taxID = staxid) |> 
#   anti_join(all.lineages) |>  pull() -> taxes
# 
# taxes |> 
#   map(insect::get_lineage, taxonomy) -> new.taxes
# new.taxes <- set_names(x = new.taxes,nm =  taxes)
# 
# new.taxes %>% discard(function(x) is.na(x[[1]]) ) ->  new.taxes
# 
# new.taxes|> 
#   map_dfr(~ bind_cols(.x) |>  mutate (rank = names(.x)) ,.id = "taxID") -> new.taxes
# 
# new.taxes |> 
#   dplyr::rename(name = 2) |> 
#   filter (rank %in%  c( "kingdom", "phylum", "class", "order", "family","genus" , "species")) |> 
#   mutate(taxID = as.numeric(taxID)) -> new.taxes
# 
# new.lineages.good |> 
#   bind_rows(lineages3)|> 
#   bind_rows(new.taxes) |> 
#   filter (rank %in%  c( "kingdom", "phylum", "class", "order", "family","genus" , "species")) |> 
#   distinct() -> all.lineages

lineages |> 
  group_by(taxID, rank) |> 
  tally() |> 
  filter (n >1) |> 
  distinct(taxID) %>% 
  anti_join(lineages,.) |> 
  pivot_wider(names_from = "rank", values_from = "name") %>%
  inner_join(output_blastp |> 
              filter (pident > 90) |>
              separate_rows(staxid, sep = ";", convert = T) |> 
              dplyr::rename(taxID = staxid),.) -> ready


custom.lca(df = ready,cutoff = 97,keep.unresolved = F) -> finalid_blastp
```
## Combine all results

```{r}
output |> distinct(qseqid) |>
  left_join(Hash.clas.last, by = c("qseqid" = "Hash")) |> 
  dplyr::rename(starting = taxa) -> starting
starting |> 
  full_join(newIDs.BOLD, by = c("qseqid" = "Hash")) |> 
  dplyr::rename(BOLD = taxa) -> with.bold

## reduce artificial precision of BLASTP and BLASTx
finalid_blastx |> 
   separate(consensus, into = taxonomy.cols, sep = "%") |> 
  mutate (genus = NA, species = NA) |> 
  unite(-qseqid, sep = "%", col = "consensus") -> finalid_blastx

finalid_blastp |> 
   separate(consensus, into = taxonomy.cols, sep = "%") |> 
  mutate (genus = NA, species = NA) |> 
  unite(-qseqid, sep = "%", col = "consensus") -> finalid_blastp


newIDs.BOLD |>
  rename(qseqid = Hash, BOLD = taxa) |> 
  full_join(finalid_blastx) |>
  rename(BLASTx = consensus) |> 
  full_join(finalid_blastp) |> 
    rename(BLASTp = consensus, Hash = qseqid) |> 
  inner_join(Hash.clas.last) -> together


```


The criteria is first look if BOLD could not get to anything, then trust BLAStx and BLASTp, only to a family level, given the lack of resolution of the AA sequence.

BLASTp has been done with many sequences. we only need to keep that in those in which the resolution of the BLAST was worse




```{r}
together |> 
   mutate(final_id = case_when(BOLD == "NA|NA|NA|NA|NA|NA|NA"   ~ BLASTx,
                              !is.na(BOLD)                      ~ BOLD,
                              !is.na(BLASTx)                    ~ BLASTx ,
                              TRUE                              ~ taxa)) -> together
                              
```

```{r}
together |> 
  filter (str_detect(final_id, "Insecta")) |> View()
  # separate (taxa, into = taxonomy.cols, sep = "\\|") |> 
  # mutate (precision = case_when(species == "NA" & genus == "NA" & family !="NA" ~ "family",
  #                               species == "NA" & genus != "NA" & family !="NA" ~ "genus",
  #                               species != "NA" & genus != "NA" & family !="NA" ~ "species")) |> 
  # select (Hash, precision) |> 
  # right_join(together)|> 
  # mutate (final_id = case_when(!is.na(final_id)   ~ final_id,
  #                            BLASTp == "NA%NA%NA%NA%NA%NA%NA"   ~ "NA%NA%NA%NA%NA%NA%NA",
  #                            TRUE                               ~ taxa)) |> 
  # filter (!is.na(final_id)) |> View()
```

